---
title: "Assignment 01 - Solutions"
subtitle: "Statistical Computing and Empirical Methods"
author: "YOUR NAME HERE"
output:
  html_document:
  toc: true
toc_depth: 2
---

*****

In this first lab session you will become familiar with the basic building
blocks of the R language. By the end of this lab you should be able to

- create and manipulate R objects
- understand the main data types used in R
- work with vectors, matrices, lists and data frames
- inspect the structure of R objects
- subset objects using indexing
- understand how R performs automatic type conversion (coercion)

Throughout this lab exercise you are encouraged to experiment with the code.
Modify values, change function arguments, and check what happens. Learning R
is much easier if you actively explore the language rather than simply copying
commands.



### Before you begin

Open **RStudio** and open the Rmarkdown document provided for this week. Unless otherwise stated, execute every code chunk and make sure you understand its output before moving on.

> **Tip**
>
> If something unexpected happens, don't panic! Read the error message
> carefully. It often tells you exactly what went wrong.

*****

# Part I: basic R 

At the R prompt, everything you type is an **expression**.

Run the following code.

```{r}
x <- 10 # using `<-` for attribution 
y = 5   # using `=` for atribution. Both work just fine.

x

z <- x + y   # attribution by itself does not print anything to the command line
x * y        # Operations without attribution print results by default
(w <- x / z) # If you enclose your attribution in parentheses it attributes and prints
```

Both `<-` and `=` assign a value to an object. 

`<-` is the older / classic attribution operator which I (Felipe) use more often - mainly because I'm **old** :). R has accepted the more standard `=` as an attribution operator for many years. Feel free to use whichever you prefer.


### Exercise 1

Create two new variables called `a` and `b`, assigning them the values 7 and 13.

Calculate

- their sum;
- their product;
- the square of `a`.

```{r ex01}
# Write your code in this code block
a = 7
b= 13
a+b
a*b
a*a


```



## Basic data types

R contains several fundamental data types.

Run the following code.

```{r}
x <- 3.14
y <- 5L
z <- "Statistics"
w <- TRUE

class(x)
class(y)
class(z)
class(w)
```

You should observe four common classes:

- `numeric`
- `integer`
- `character`
- `logical`

These classes form the basis of almost every object you will use later in the
course.

## Special numerical values

Run the following commands.

```{r}
1 / 0

0 / 0

sqrt(-1)
```

Notice the appearance of special values such as

- `Inf`
- `NaN`

These represent mathematical quantities that cannot be expressed as ordinary
numbers.

### Exercise 2

Predict the class of each object before running the code.

```{r ex02}

class(FALSE) #logical

class("123") #character

class(123) #numeric

class(123L) #integer
```

Were your predictions correct?


## Vectors

Vectors are probably the most fundamental data structure in R.

```{r}
temperatures <- c(18.4, 19.2, 21.1, 20.3, 22.0)

temperatures
```

Check its `class` and `length` attributes.

```{r}
class(temperatures)

length(temperatures)
```

Individual vector elements can be extracted using square brackets.

```{r}
temperatures[1]

temperatures[3]

temperatures[c(2,4)] # you can pass a vector of indices

temperatures[2:5]    # a:b produces an integer sequence from a to b.
```

**Important**: Remember that **R starts indexing at 1**, not 0. Try typing `temperatures[0]` to see what happens.


### Exercise 3

Create a vector containing the numbers

```
2 4 6 8 10 12
```

using whichever method you prefer. Store it in an object called `even_numbers` and print that object to the console.

```{r ex03a}
# Write your code in this code block
even_numbers = c(2, 4, 6, 8, 10, 12)
even_numbers

```

Using indexing only,

1. extract the first element;
2. extract the last element;
3. extract the middle two elements.

```{r ex03b}
# Write your code in this code block
even_numbers[1]
even_numbers[6]
even_numbers[c(3,4)]
```

Replace the fourth element of your vector by the value `100` and print the resulting vector.

```{r ex03c}
# Write your code in this code block
even_numbers[4]=100
even_numbers

```

## Automatic coercion

Unlike many programming languages, R automatically converts data types whenever
it thinks this is appropriate. Run the following examples.

```{r}
2 + TRUE

FALSE + FALSE

TRUE + TRUE
```

Can you explain the output?


Now try

```{r}
c("A", "B", 5)

c(TRUE, 1, FALSE)

c(3.2, 7L)
```

Notice that all elements of a vector must belong to the same class. When different types are combined, R will try to coerce the values to the most general data type that still keeps them consistent.

### Exercise 4

Without running the code first, predict what each command will return.

```{r}
c(1, TRUE) #1 1 

c("1", TRUE) # 1 TRUE

c(5L, 2.5) # 5.0 2.5
```

Now execute the commands. Were your predictions correct?


## Inspecting objects

As objects become more complicated, simply printing them is often insufficient. The function `str()` can be quite useful in these cases. Try running:

```{r}
student <- list(name = "Alice",
                age = 23,
                passed = TRUE)

student

str(student)
```

`str()` displays the internal structure of the object rather than just printing its contents.

### Exercise 5 (challenge)

Create a list describing yourself. Include at least

- your first name;
- your age;
- your favourite programming language;
- whether you have previously used R.

Store the result in an object called `mylist`. Then display its structure using `str()`.

```{r ex05}
# Write your code in this code block
mylist = list (name = "Nicole",
                age = 21,
                'favourite programming language'= "Python",
                'previously used R' = FALSE)

str(mylist)

```

## Missing values

Missing observations are represented by the special value `NA`. Run the following code:

```{r}
scores <- c(74, 68, NA, 91, 82)

scores
```

You can check which elements are missing with:

```{r}
is.na(scores)
```

### Exercise 6

What happens if you calculate the following expression?

```{r}
mean(scores)
```

Why does this happen? Can you find by either checking the function documentation, `?mean`, or your own experimentation, a way to calculate the mean while ignoring missing values?

```{r ex06}
mean (scores [c(1,2,4,5)])

```

# Exercise 7

Suppose a researcher records the number of hours studied by six students.

```{r}
hours <- c(5, 7, 4, 8, NA, 6)
```

Without creating another vector and using only R functions, do the following:

1. determine how many observations are missing;
2. replace the missing value with 5;
3. calculate the average number of study hours;
4. replace every value greater than 6 by exactly 6.

```{r ex07}
# Write your code in this code block
sum(is.na(hours))
hours[is.na(hours)]=5
mean(hours)
hours[hours>6]=6

```

*****

# Part II: R data structures

## Matrices

Matrices are essentially vectors with an additional **dimension** attribute. Create the following matrix.

```{r}
A <- matrix(1:12,
            nrow = 3,
            ncol = 4)

A
```

Notice that the numbers are filled **column by column**. You can check its dimensions with:

```{r}
dim(A)
nrow(A)
ncol(A)
```

Accessing individual matrix values requires **two indices**, with the notation `[row, column]`:

```{r}
A[1, 1] # first row, first col
A[3, 2] # third row, second col
A[, 4]  # a blank index means "all elements along this dimension"
A[2, ]
A[1:2, c(1,4)] # you can also use vectors to index multiple elements
```


### Exercise 8

Create the following matrix M using `matrix()`.

$$
M = \begin{bmatrix}
1 & 2 & 3\\
4 & 5 & 6\\
7 & 8 & 9
\end{bmatrix}
$$

Note that it will look differently in the R console. Your code will produce something that looks like this:

```{r, eval=FALSE}
[,1] [,2] [,3]
[1,]    1    2    3
[2,]    4    5    6
[3,]    7    8    9
```

**Hint**: look at the `byrow` argument of `matrix()` (use `?matrix` to access the function documentation).

```{r ex08a}
# Write your code in this code block
A <- matrix(1:9,
            nrow = 3,
            ncol = 3,,
            byrow=TRUE)

A

```

Using indexing only,

- extract the entire second row;
- extract the entire first column;
- extract the bottom-right element;
- replace the centre element by 99.

```{r ex08b}
# Write your code in this code block
A[1, ]
A[ , 1]
A[3,3]
A[2,2]=99

```

*****

You can inspect the attributes of a matrix using

```{r}
attributes(A)
```

Check what attributes a matrix possesses that a vector does not?


## Lists

Unlike vectors and matrices, **lists can contain objects of different types**. Run the following code:

```{r}
experiment <- list(researcher = "Alice",
                   participants = 25,
                   completed = TRUE,
                   measurements = c(12.4, 14.2, 11.9))

experiment
```

Now inspect its structure,

```{r}
str(experiment)
```

## Accessing list elements

List elements can be accessed either by **name**

```{r}
experiment$researcher

experiment$measurements
```

or by **position**, using double square brackets:

```{r}
experiment[[1]]

experiment[[4]]
```

### Exercise 9

Create a list describing one of your favourite films (or books). Include at least

- title;
- year;
- genre;
- rating out of 10;
- three main actors (or characters).

Store the result in an object called `movie`.

```{r ex09a}
# Write your code in this code block
movie <- list(title = "A song of ice and fire",
                   year = 2005,
                   genre = "fantasy",
                   'rating out of 10' = 9,
                   'three main characters'="Daenarys Targaryen, Eddard Stark, Catelyn Stark")

movie

```

Using your object `movie`,

- print only the title;
- print only the rating;
- change the rating;
- add a new element called `watched` with value `TRUE`.

```{r ex09b}
# Write your code in this code block
movie ["title"]
movie["rating out of 10"]
movie["rating out of 10"]=8
movie["watched"]=TRUE

```

What is the difference between the following two commands:

```{r}
movie[1]
```

and

```{r}
movie[[1]]
```

Try both commands, attribute each to one variable (use any name you prefer) and inspect the results with `class()`. Write one or two sentences explaining the difference.

> Answer:
> the movie [[1]] only gives the infromation about the element itself, wehreas movie [1] gives more information on the type of element it is in the list
> 


## Factors

Many variables encountered in statistics are **categorical** rather than
numerical. Examples include

- country;
- blood type;
- species;
- treatment group;
- etc.

In R these variables are usually stored as **factors**. Check the following:

```{r}
group <- c("Treatment",
           "Treatment",
           "Control",
           "Treatment",
           "Control")

group
```

This is a simple character vector. To convert it into a factor, do:

```{r}
group_factor <- factor(group)

group_factor
```

Inspect its structure.

```{r}
str(group_factor)

levels(group_factor)
```

Although character vectors can also store text, factors provide R with
additional information:

- the set of valid values (levels);
- whether the variable is categorical;
- the order of categories (for ordered factors).

Many statistical modelling functions rely on factor variables.

### Exercise 10

Create a factor representing the weekdays

```
Monday
Tuesday
Wednesday
Thursday
Friday
```

Store it in an object called `weekdays` and display its levels using `levels()`.

```{r ex10a}
# Write your code in this code block

```

Suppose a variable can take the following values:

```
Low
Medium
High
```

Create a factor with these values in the correct order.

Hint: look at the description of the `levels` argument in `?factor`.

```{r ex10b}
# Write your code in this code block

```

# Data frames

Most datasets used in data science are stored as **data frames**: 

- Each row represents one observation;
- Each column represents one variable.

We can create a small data frame like this:

```{r}
students <- data.frame(Name = c("Alice", "Bob", "Carol", "David"),
                       Age = c(22, 24, 23, 21),
                       MSc = c(TRUE, TRUE, FALSE, TRUE))

students
```

Inspect the object.

```{r}
class(students)

str(students)

names(students)
```

Columns may be accessed in several ways.

```{r}
students$Age

students[["Age"]]

students[, 2]
```

Rows can also be extracted.

```{r}
students[2, ]

students[1:3, ]
```

### Exercise 11

Add a new column called `Mark` to the dataframe above. Choose any four marks between 50 and 100 for the example students.

```{r ex11a}
# Write your code in this code block

```

Now calculate

- the mean mark;
- the maximum mark;
- the minimum mark.

```{r ex11b}
# Write your code in this code block

```

Extract only the students whose mark is greater than 70.

**Hint**: `students$Mark > 70` returns a logical vector that can be used for indexing.

```{r ex11c}
# Write your code in this code block

```



## Names and attributes

Many R objects carry additional information called **attributes**. Inspect the attributes of your data frame:

```{r}
attributes(students)
```

Notice the presence of

- row names;
- column names;
- class.

### Exercise 12

Rename the column `Mark` to `FinalMark`. **Hint**: `names(students)` can both display and be used to modify column names by attribution.

```{r ex12}
# Write your code in this code block

```

### Exercise 13 (challenge)

Create a data frame describing five countries. Include the variables

- Country
- Population (millions)
- Continent
- EU_member (TRUE/FALSE)

```{r ex13a}
# Write your code in this code block

```

Then answer the following questions using R:

1. Which country has the largest population?
2. How many countries are EU members?
3. What is the average population?
4. Display only the countries not members of the EU.

```{r ex13b}
# Write your code in this code block

```

*****


# Part III: Functions, conditionals and loops

## Calling functions

You have already used many R functions without thinking about them. For example,

```{r}
x <- c(3, 8, 5, 10, 12)

length(x) # function `length()`
mean(x)   
sort(x)   
summary(x)
```

Each function receives zero or more inputs (arguments), performs some computation, and returns some output. 

Different functions accept different numbers of arguments - some required, some optional. For example,

```{r}
rnorm(5)

rnorm(5, mean = 10)

rnorm(mean = 10, n = 5, sd = 2)
```

Notice that arguments can be supplied by position or by name. Using argument names generally makes your code easier to read and less susceptible to hard-to-find bugs.

### Exercise 14  

Read the help page

```r
?runif
```

Now generate:

- 20 random numbers between 0 and 1;
- 20 random numbers between 5 and 10;
- 100 random numbers between -2 and 2.

```{r ex14}
# Write your code in this code block

```

## Writing your first function

Suppose that we frequently need to calculate the area of a rectangle. Rather than repeatedly writing:

```r
width * height
```

we can define a function.

```{r}
rectangle_area <- function(width, height){
  area <- width * height
  return(area)
}
```

You can then call the function whenever it's needed:

```{r}
rectangle_area(4, 5)
rectangle_area(width  = 10, height = 3)
```

Notice the general structure.

```r
name <- function(arguments){

commands

return(result)

}
```

### Exercise 15

Write a function called `circle_area` that receives a radius and returns the corresponding area of the circle. Test your function using `radius = 1` and `radius = 5`.

```{r ex15a}
# Write your code in this code block

```

Write a function called `fahrenheit_to_celsius` that converts temperatures from Fahrenheit to Celsius (do a quick internet search if you can't recall the formula). Test it on

- 32$^\circ$ F
- 68$^\circ$F
- 100$^\circ$F

```{r ex15b}
# Write your code in this code block

```

## Default argument values

Function arguments can have default values. Consider the following function definition:

```{r}
roll_die <- function(nrolls = 1,
                     nsides = 6){
  sample.int(nsides,
             size = nrolls,
             replace = TRUE)
}
```

Now try:

```{r}
roll_die() # single roll of a 6-sided die
roll_die(5) # 5 rolls of 6-sided dice. 
roll_die(nsides = 20) # 1 roll of a 20-sided die
roll_die(10, 20) # 10 rolls of a 20-sided dice
```

### Exercise 16

Modify your `circle_area()` function so that the radius defaults to 1. Check what happens when you run `circle_area()`

```{r ex16}
# Write your code in this code block

```

## Decisions using `if`

Programs often need to make conditional decisions. The simplest decision structure in R is

```r
if(condition){

commands

} else {

other commands

}
```

Run the following example.

```{r}
temperature <- 12

if(temperature < 15){
  print("It is cold today.")
} else {
  print("No coat needed.")
}
```

Notice that defining an `else {...}` is optional.


### Exercise 17

Write a function called `pass_fail` that receives a mark and returns `"Pass"` if the mark is at least 50;or returns `"Fail"` otherwise.

Test your function using the values `40`, `50` and `70`.

```{r ex17}
# Write your code in this code block

```


## Vectorised decisions with `ifelse()`

The ordinary `if` statement only works with a **single** logical condition. Suppose instead that we have a vector.

```{r}
marks <- c(45, 61, 72, 38, 55, 81)
```

To classify every observation simultaneously we can use

```{r}
ifelse(test = marks >= 50,
       yes  = "Pass",
       no   = "Fail")
```

Notice that one command processes the entire vector.

### Exercise 18

Given the vector

```{r}
ages <- c(12, 15, 18, 21, 34, 9)
```

use `ifelse()` to create a character vector containing `"Adult"` for ages 18 or above and `"Minor"` otherwise.

```{r ex18}
# Write your code in this code block

```

## Repeating computations with `for`

Suppose we wish to print the first five squares. We can use the loop function `for()`:

```{r}
for(i in 1:5){
  print(i^2)
}
```
Loops become particularly useful when generating and storing results. For instance (note: there's a simpler way of doing the operation below. We'll discuss it later):

Loops become particularly useful when storing results.

```{r}
# Create some example data
scores <- data.frame(maths   = rnorm(5, 70, 8),
                     english = rnorm(5, 68, 10),
                     science = rnorm(5, 72, 7))

# Create an empty numeric vector to store the means
score_means <- numeric(ncol(scores))

# Loop over the columns
for(i in 1:ncol(scores)) {
  score_means[i] <- mean(scores[[i]])
}

scores
score_means
```

Notice that we created the output vector **before** entering the loop. This is called **preallocation**, and it is considerably faster than growing objects one element at a time.

### Exercise 19

Write a `for` loop to calculate the _range_ (difference between the maximum and the minimum value) of each column of the `scores` dataframe, and store the results in a vector called `ranges`.

```{r ex19}
# Write your code in this code block

```

## `while` loops

Unlike a `for` loop, a `while` loop continues until a condition becomes `FALSE`. Try the following:

```{r}
x <- 1

while(x < 100){
  x <- x * 2
  print(x)
}
```

How many iterations were required?


## Using functions inside loops

The real power of programming comes from combining functions and loops. For instance, the following code generates a fractal sequence known as the _Sierpi\'nski triangle_:

```{r}
# function to randomly choose one vertex of a triangle
choose_vertex <- function(vertices) { vertices[sample(1:3, 1), ] }

# MAtrix representing a triangle with sides of length 1
vertices <- matrix(c(0, 0,
                     1, 0,
                     0.5, sqrt(3)/2), 
                   ncol = 2, byrow = TRUE)

# Empty matrix to receive the results of the computations
points <- matrix(nrow = 10000, ncol = 2)

# Starting point (feel free to change it)
points[1, ] <- c(0.2, 0.2)

for(i in 2:nrow(points)) {
  # populate next entry with a point halfway between the previous one and 
  # a randomly selected vertex
  vertex <- choose_vertex(vertices)
  points[i, ] <- (points[i - 1, ] + vertex) / 2
}

# Plot the results
plot(points, pch = ".", asp = 1)
```



### Exercise 20 (chalenge)

Modify the _Sierpi\'nski triangle_ code above so that each point is coloured according to its distance from the three vertices of the triangle. Specifically:

1. Inside the loop, compute the distance from each generated point to each of the three vertices. 
- **Hint**: you may need another `for` loop inside the original one to calculate the distance between `points[i, ]` and each vertex. 
- **Hint**: You can calculate the Euclidean distance between two vectors using `sqrt(sum((vector_1 - vector_2)^2))`
2. Use these three distance values as the red, green, and blue (RGB) colour components. The function `rgb()` creates colours from red, green, and blue values between 0 and 1 (check `?rgb`). Store the colour valur for each point in a vector `my_cols`
3. Plot the points using these colours (*Hint**: function `plot()` accepts a colour argument using `col = ...`)

```{r ex20}
# Write your code in this code block

```

## `apply()`

`apply()` is designed for **matrices** (and higher-dimensional arrays). Instead of using a `for` loop, we can compute functions on each row/column simply by doing:

```{r}
# Example matrix
m <- matrix(1:20, nrow = 4)

# Mean of each column
apply(m, MARGIN = 2, FUN = mean)

# Mean of each row
apply(m, MARGIN = 1, FUN = mean)
```

### Exercise 21

Use `apply()` to compute

1. the maximum (`max()`) of every row;

2. the standard deviation (`std()`) of every column;


```{r ex21}
# Write your code in this code block

```

# `lapply()`

`apply()` works on matrices. `lapply()` has a similar role, but it works on **lists** (also data frames, since they are just specialised lists).

```{r}
numbers <- list(a = 1:5,
                b = 1:10,
                c = 1:20)

# mean of every list element.
lapply(numbers, mean)
```

Notice that the output is itself a **list**.

---

### Exercise 22

Use `lapply()` to compute the length and the sum of each element of the list `numbers`.

```{r ex22}
# Write your code in this code block

```

# The `map()` family

The tidyverse provides a modern alternative to the `*apply()` functions through the **purrr** package.

Load the package.

```{r}
# install.packages("purr") # Uncomment and install if needed
library(purrr)
```

The basic function is `map()`:

```{r}
map(numbers, mean)
```

Like `lapply()`, this returns a list. One of the advantages of the `map()` over the `*apply()` family is that rather than trying to infer how to simplify its output automatically (which is error-prone), **purrr** provides explicit functions for different output types.

```{r}
# Always returns a vector of doubles
map_dbl(numbers, mean)
# Always returns a vector of integers
map_int(numbers, length)
```

Because the output type is known in advance, these functions tend to make
programs easier to understand and debug.

---

### Exercise 23

Use `map_dbl()` to compute the maximum of each list element, and `map_int()` to compute the length of each element.

```{r ex23}
# Write your code in this code block

```

## Anonymous functions

We are not restricted to built-in functions. We can define a function on the fly, as needed. For instance,

```{r}
map_dbl(numbers,
        \(x) sum(x ^ 2))
```

Here the anonymous function computes the sum of squares of each vector.

### Exercise 24 (challenge)

Use `map_dbl()` with anonymous functions to compute:

1. the average of the squared values;
2. the number of even integers;
3. the proportion of values greater than 5.

```{r ex24}
# Write your code in this code block

```

---

Suppose we want the average of each column of a matrix

```{r}
m <- matrix(rnorm(25), nrow = 5)
```

Now compare the following (functionally equivalent) solutions:

```{r}
# ==== version 1 ====
means <- numeric(ncol(m))

for(i in seq_len(ncol(m))){
  means[i] <- mean(m[, i])
}

means


# ==== version 2 ====
apply(m, 2, mean)
```


Which version do you find easier to read? However, keep in mind that you cannot use `apply()` or `map()` to replace all loops - when iterations are not independent (i.e., when the result of a given iteration depends on the previous ones) then you need to use the explicit loops.


## Exercise 25 (challenge)

The built-in dataset `iris` contains measurements on three flower species.

```{r}
head(iris)
```

Using **only** iteration functions (`apply()`, `lapply()`, `sapply()` or
`map()`), calculate

1. the mean of each numeric variable;

2. the standard deviation of each numeric variable;

3. the number of missing values in each column.

**Hint**: `iris[, 1:4]` extracts the numeric columns.

**Hint**: Function `is.na()` and arguments `na.rm` of the mean and standard deviation functions may be useful.

```{r ex25}
# Write your code in this code block

```
